Supplementary Materials: Separating random and deterministic sources of computational noises in explore-exploit decisions

### 2024-05-16 manuscript/2024 - Plot Comp Bio/manuscript_DeterministicRandomNoise.tex
final version 240515

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/RanDetNoise_modelfree.jpg}
			\caption{Replication of previous findings that people use both random and directed exploration in this task. (A) model-free measure of behavioral variability, $p(\mbox{low mean})$, increases with horizon. (B) model-free measure of information seeking, $p(\mbox{high info})$, increases with horizon. (C) model-based measure of behavioral variability, decision noise $\sigma$, increases with horizon. (B) model-based measure of information seeking, information bonus $A$, increases with horizon.}
			\label{fig:modelfree}
		\end{center}
	\end{figure}

@@

\subsubsection*{Model validation}
To be sure that our fit parameter values were meaningful and to understand the limits of our model, we evaluated our model extensively using simulated data. This allowed us to quantify whether deterministic and random noise can be identified under ideal conditions where the behavior is generated by the model with known parameters.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/RDBayes_parameterrecovery_subject_examplesession_randet.jpg}
		\caption{Parameter recovery analysis for random (A,B) and deterministic noise (C,D) in the two horizons.}
		\label{fig:paramrecover_main}
	\end{center}
\end{figure}

Parameter recover in this task was good for this model, with fit values for $\sigma_{ran}$ and $\sigma_{det}$ showing strong correlations with their simulated values in both horizon (H) conditions (For $\sigma_{ran}$, $R = 0.91 (\mbox{H} = 1) \mbox{ and } 0.84 (\mbox{H} = 6), p < 0.001$, For $\sigma_{det}$, $R = 0.60 (\mbox{H} = 1) \mbox{ and } 0.59 (\mbox{H} = 6), p < 0.001$). However, while the relationship was near perfect for random noise ($\frac{\mbox{\small Recovered } \sigma_{ran}}{\mbox{\small Simulated } \sigma_{ran}} = 1.01$), there was a systematic bias to underestimate the level of deterministic noise by about 32\% ($\frac{\mbox{\small Recovered } \sigma_{det}}{\mbox{\small Simulated } \sigma_{det}} = 0.68$). Despite this underestimation of deterministic noise in both horizon conditions, the difference in deterministic noise between horizons is much better captured (See Supplementary Materials section 2.2). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizon conditions. In addition, we see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games. 
	% $\frac{\mbox{\small Recovered } \Delta \sigma_{det}}{\mbox{\small Simulated } \Delta \sigma_{det}} = 0.74$

In addition to the conventional subject-level parameter recovery analysis presented here, we also performed parameter recovery analysis that examined how faithful the full posterior distribution of group-level parameters can be recovered in simulated data (Supplementary Figure S5, S6). Qualitatively, we also showed that our way of modeling deterministic noise is capable of capturing known deterministic processes intentionally omitted from the full model (Supplementary Figure S4). Full details of these additional analyses are presented in the Supplementary Materials.

@@

\begin{figure}[H]
\begin{center}
	\includegraphics[width=0.7\textwidth]{figures/RDBayes_hyperprior.jpg}
	\caption{Model based analysis showing the posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and change with horizon (B, D).  }
	%BOB SAYS: However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.
	\label{fig:mb1}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
	\includegraphics[width=0.5\textwidth]{figures/RDBayes_ratio__paironly.jpg}
	\caption{Model based analysis showing the posterior distributions over the ratio of the group-level mean of the standard deviations of  random and deterministic noise between horizon 6 and horizon 1 respectivelly. The ratio in the standard deviations of noises between horizon 6 and horizon 1 is similar for random and deterministic noise.}
	\label{fig:ratio}
\end{center}
\end{figure}

@@

As a control, we also applied posterior predictive checks on alternative models that consider only deterministic or only random noise, and these models fail to capture all qualitative patterns (Supplementary Figure S13). Full details of this analysis can be found in Supplementary Materials section 3.2.

\begin{figure}[H]
\begin{center}
	\includegraphics[width=1\textwidth]{figures/RDBayes_2noise_modelA.jpg}
	\caption{
		Our model accounts for all qualitative patterns of the data, namely, p(high info) and p(low mean) increase as a function of horizon, p(inconsistent) increases as a function of horizon for both [1 3] and [2 2] conditions and it lies between the pure random and pure deterministic noise prediction.}
	\label{fig:mb3}
\end{center}
\end{figure}

>>>>

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/RanDetNoise_modelfree.jpg}
			\caption{Replication of previous findings that people use both random and directed exploration in this task. (A) model-free measure of behavioral variability, $p(\mbox{low mean})$, increases with horizon. (B) model-free measure of information seeking, $p(\mbox{high info})$, increases with horizon. (C) model-based measure of behavioral variability, decision noise $\sigma$, increases with horizon. (D) model-based measure of information seeking, information bonus $A$, increases with horizon.}
			\label{fig:modelfree}
		\end{center}
	\end{figure}

@@

\subsubsection*{Model validation}
To be sure that our fit parameter values were meaningful and to understand the limits of our model, we evaluated our model extensively using simulated data. This allowed us to quantify whether deterministic and random noise can be identified under ideal conditions where the behavior is generated by the model with known parameters. Full details of this analysis is presented in the Supplementary Materials section 2.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/RDBayes_parameterrecovery_subject_examplesession_randet.jpg}
		\caption{Parameter recovery analysis for random (A,B) and deterministic (C,D) noises in the two horizons.}
		\label{fig:paramrecover_main}
	\end{center}
\end{figure}

Parameter recover in this task was good for this model, with fit values for $\sigma_{ran}$ and $\sigma_{det}$ showing strong correlations with their simulated values in both horizon (H) conditions (For $\sigma_{ran}$, $R = 0.91 (\mbox{H} = 1) \mbox{ and } 0.84 (\mbox{H} = 6), p < 0.001$, For $\sigma_{det}$, $R = 0.60 (\mbox{H} = 1) \mbox{ and } 0.59 (\mbox{H} = 6), p < 0.001$). However, while the relationship was near perfect for random noise ($\frac{\mbox{\small Recovered } \sigma_{ran}}{\mbox{\small Simulated } \sigma_{ran}} = 1.01$), there was a systematic bias to underestimate the level of deterministic noise by about 32\% ($\frac{\mbox{\small Recovered } \sigma_{det}}{\mbox{\small Simulated } \sigma_{det}} = 0.68$). Despite this underestimation of deterministic noise in both horizon conditions, the difference in deterministic noise between horizons is much better captured (see Supplementary Materials section 2.2). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizon conditions. In addition, we see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games. 
	% $\frac{\mbox{\small Recovered } \Delta \sigma_{det}}{\mbox{\small Simulated } \Delta \sigma_{det}} = 0.74$

In addition to the conventional subject-level parameter recovery analysis presented here, we also performed parameter recovery analysis that examined how faithful the full posterior distribution of group-level parameters can be recovered in simulated data (Supplementary Figure S5, S6). Qualitatively, we also showed that our way of modeling deterministic noise is capable of capturing known deterministic processes intentionally omitted from the full model (Supplementary Figure S4). Full details of these additional analysis are presented in the Supplementary Materials.

@@

\begin{figure}[H]
\begin{center}
	\includegraphics[width=0.7\textwidth]{figures/RDBayes_hyperprior.jpg}
	\caption{Model based analysis showing the posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and increase with horizon (B, D).  }
	%BOB SAYS: However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.
	\label{fig:mb1}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
	\includegraphics[width=0.5\textwidth]{figures/RDBayes_ratio__paironly.jpg}
	\caption{(A) Model based analysis showing the posterior distributions over the ratio of the group-level mean of the standard deviations of  random and deterministic noise between horizon 6 and horizon 1 respectivelly. The ratios in the standard deviations of noises between horizon 6 and horizon 1 are similar for random and deterministic noise.}
	\label{fig:ratio}
\end{center}
\end{figure}

@@

As a control, we also applied posterior predictive checks on alternative models that consider only deterministic or only random noise, and these reduced models fail to capture all qualitative patterns (Supplementary Figure S13). Full details of this analysis can be found in Supplementary Materials section 3.2.

\begin{figure}[H]
\begin{center}
	\includegraphics[width=1\textwidth]{figures/RDBayes_2noise_modelA.jpg}
	\caption{
		Our model accounts for all qualitative patterns of the data, namely, (A) p(high info) and (B) p(low mean) increase as a function of horizon, p(inconsistent) increases as a function of horizon for both [1 3] (C) and [2 2] (D) conditions and it lies between the pure random and pure deterministic noise prediction.}
	\label{fig:mb3}
\end{center}
\end{figure}

### 2024-05-16 manuscript/2024 - Plot Comp Bio/supplemental/DetRanNoise_Supplementary.tex
final version 240515

	\section{Additional model-free analyses}
	\subsection{Replication of Figure 2 without excluding subjects}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/RanDetNoise_modelfree__all.jpg}
			\caption{Replication of previous findings with data from all participants (i.e. no exclusions). Both  $p(\mbox{low mean})$ (A) and $p(\mbox{high info})$ (B) increase with horizon suggesting that people use both random and directed exploration in this task.  }
			\label{fig:s1}
		\end{center}
	\end{figure}
	\newpage

@@

	\section{Additional model-based analyses}
	\subsection{Replication of Figure 5 without excluding subjects}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/RDBayes_hyperprior__all.jpg}
			\caption{Model based analysis with data from all participants (i.e. no exclusions) showing the posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and change with horizon (B, D).  However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.}
			\label{fig:s8}
		\end{center}
	\end{figure}
	\newpage
	\subsection{Alternative variations of the two-noise model}
	To check whether all aspects of the model were necessary to reproduce the qualitative pattern of findings, we also built and fit five additional versions of the model. These models varied in whether deterministic and random noise are present or not and whether either types of noise is dependent on horizon. Specifically, we tested the following 6 models (Note that the $\sigma^{ran}_{horizon},\sigma^{det}_{horizon}$ model is our original full model).

>>>>

	\section{Additional model-free analyses}
	\subsection{Replication of Figure 2 without excluding subjects}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/RanDetNoise_modelfree__all.jpg}
			\caption{Replication of previous findings with data from all participants (i.e. no exclusions).  (A) model-free measure of behavioral variability, $p(\mbox{low mean})$, increases with horizon. (B) model-free measure of information seeking, $p(\mbox{high info})$, increases with horizon. (C) model-based measure of behavioral variability, decision noise $\sigma$, increases with horizon. (D) model-based measure of information seeking, information bonus $A$, increases with horizon.}
			\label{fig:s1}
		\end{center}
	\end{figure}
	\newpage

@@

	\section{Additional model-based analyses}
	\subsection{Replication of Figure 5 without excluding subjects}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/RDBayes_hyperprior__all.jpg}
			\caption{Model based analysis with data from all participants (i.e. no exclusions) showing the posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and change with horizon (B, D).  However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.}
			\label{fig:s8}
		\end{center}
	\end{figure}
	\newpage
	\subsection{Alternative models: variations of the two-noise model}
	To check whether all aspects of the model were necessary to reproduce the qualitative pattern of findings, we also built and fit five additional versions of the model. These models varied in whether deterministic and random noise are present or not and whether either types of noise is dependent on horizon. Specifically, we tested the following 6 models (Note that the $\sigma^{ran}_{horizon},\sigma^{det}_{horizon}$ model is our original full model).
